Rationality, Evolution \& the Environment

### 2015-06-10 paper/paper.tex
Paolo new version

\usepackage{a4wide}

@@

The standard use of the word \textit{rationality} in economics and decision theory is to denote the maximization of subjective expected utility. Since the work by vNM [cite], this stance has been justified from the intuitive ''rationality'' of the axioms on the subjective preference relation leading to the representability of the decision maker as a subjective expected utility (SEU) maximizer. 
While in [vNM] the probability of the outcomes is objectively given a priori, Savage54 provides a set of axioms that represent the decision maker (DM) with both a subjective utility over outcomes \textit{and} a \textit{subjective} probability over states of the world. Savage's axioms constitute the foundation of subjective expected utility theory, that is normally assumed to define the rational behavior of economic agents.

The subjectivity of preferences plays a very important
role. From an agent's standpoint, the amount of, say, money achieved does not necessarily
coincide with the amount of subjective utility attained. The first to notice and formally argue for this difference
between objective utility (goods, money, etc.) and subjective utility (preferences) was
D. Bernoulli in the famous St. Petersburg paradox. Moreover, Bernoulli introduced the idea that a good representation of
subjective utility can be carried out in terms of an increasing concave function of objective
utility. This can be considered the birth of modern decision theory.

However, many violations of Savage's (and vNM's) axioms have been consistently observed in behavioral economics. Most of the attention has been focused on the Independence Axiom and the Sure-thing Principle. Allais' paradox and Ellsberg's paradox are examples of violations of these two axioms, and they are at the origin of Prospect theory [cite] and ambiguity in decision theory [cite].\\

The main idea behind Ellesberg's paradox and the literature on ambiguity is that DM treats differently objective and subjective uncertainty. If the probability measure on the possible states of the world is not given a priori and ''known'', then observed preferences leave no room to represent  DM's beliefs via a probability measure on the possible states. The most common solution is then to represent DM's beliefs with a set of probability measures instead of a single one.

Prospect theory was introduced by Kahnemann and Tversky in
their famous paper from 1979. By means of experimental results they noticed that the decision
maker apparently acts as a risk averse agent with respect to gains and as a risk loving agent
with respect to losses. Consequently, they conjectured that the subjective utility is better
representable by an S-shaped utility function, concave in the positive domain of gains and
convex in the negative domain of losses.
\\

More recently, other possible definitions of the subjective utility function have been proposed
and justified from a game theoretical point of view (e.g. Fehr and Schmidt, Charness and Rabin,
etc.). The central goal of these studies is to explain/describe the behavior in social dilemmas
(e.g., Prisoner's dilemma, public goods game, etc.). In the same spirit of Prospect theory, by means of different subjective utility
functions it is possible to describe/match the behavior observed in game theoretical experiments and to
explain it as rational (in the sense of maximizing the subjective utility).

With this approach every behavior, no matter how far from the maximization of
objective utility, can be deemed rational if we can only find a subjective utility that
justifies that behavior as maximizing the subjective utility. Quoting Hegel, (almost)
everything real is rational (this is just for fun, I will take it away, no problem). The problem here is that it seems possible to consider rational everything
that is observed, given the adequate subjective utility. The notion
of rationality almost collapses onto the observed behavior.

On the other side, other economists (Alchian, Friedman) seem to link the notion of rational choice
more tightly to the maximization of objective utility. In their classic works they argue that
profit maximization is a reasonable assumption for characterizing outcomes in a competitive
market because only firms behaving in a manner that is consistent with profit maximization will
survive in the long run. If the previous accounts of rationality looked more descriptive, this
definitely tastes more normative. In this sense, the normative notion of rationality is tightly
linked to (and almost coincide with) the maximization of objective utility, which is supposed
to drive the evolutionary process. It follows then that being normative in this sense points to an identification of rationality with maximization of objective utility.

@@

There is another violation of Savage's axioms that is particularly relevant for the present work: transitivity of preferences. The first of Savage's axioms states that the preference relation is a (total) preorder, i.e., reflexive and transitive. Moreover, the preorder axiom is often referred to as the rationality axiom [cite], as to suggest that it is somehow the core of rational preferences. Violations of transitivity have been observed and studied in experimental literature [citeTversky69,etc], but can also derive from a more theoretical and normative point of view. \\
In what follows, an important focus of our work will be the role of regret in decision making. Regret theory predicts the violation of both transitivity and Independence of Irrelevant Alternatives (IIA). [CONNECTING TO Fawcett\&al: this is IIA, not independence]

@@

Regret theory has been introduced in decision theory by Savage [] and Niherens [], and later developed by Bell [1982], Loomes and Sugden [1982], and Fishburn [1982] independently. Lately, Halpern and Pass show how the use of regret minimization can solve games like the Centipede and the Traveler's dilemma in a way closer to everyday intuition and empirical data. In this paper the notion of regret is defined as in [HalPass12]:

@@

EXAMPLE. Traveler's dilemma maybe? or maybe 2

The meta-game perspective that we introduce here can shed more light on the duality of descriptivism and normativism, and opens
some space in between the two extrema. 
By keeping the evolutionary approach we want to retain a
normative perspective on rationality, but our results show that (under certain conditions)
evolution does not necessarily lead to identifying rationality with maximization of objective
utility.

@@

The justification for this choice is both historical and behavioral. Among
the possible competitors of utility maximization, regret has always been in a prominent
position, both from a descriptive and normative viewpoint. Indeed, Bell82 writes
\begin{quote}
  By explicitly incorporating regret, expected utility theory not only becomes a better
  descriptive predictor but also may become a more convincing guide for prescribing behavior to
  decision makers.
\end{quote}
\citet{LoomesSugden1982:Regret-Theory:-} show that a regret-based theory of choice can match
and explain all the main experimental results of prospect theory, and claim that
\begin{quote} [...] we shall challenge the idea that the conventional axioms [of decision
  theory] constitute the only acceptable basis for rational choice under uncertainty. We shall
  argue that it is no less rational to act in accordance with regret theory, and that
  conventional expected utility theory therefore represents an unnecessarily restrictive notion
  of rationality. \hfill \citep{LoomesSugden1982:Regret-Theory:-}
\end{quote}

>>>>

\usepackage{bigints}
\usepackage{MnSymbol}
\usepackage{a4wide}

@@

The standard use of the word \textit{rationality} in economics and decision theory is to denote the maximization of (subjective) expected utility. Since the work by vNM [cite], this stance has been justified from the intuitive ''rationality'' of the axioms on the subjective preference relation leading to the representability of the decision maker as an expected utility maximizer. 
While in the axiomatization by [vNM] the probability of the outcomes is objectively given a priori, Savage54 provides a larger set of axioms that represent the decision maker (DM) with both a subjective utility over outcomes \textit{and} a \textit{subjective} probability over states of the world. Savage's axioms constitute the foundation of subjective expected utility (SEU) theory, that is normally assumed to define the rational behavior of economic agents.\\

DEFINITION(SEU-maximization): Let $S$ be a set of states of the world, $X$ a set of outcomes and $A$ a set of actions. Given a probability measure $P$ on $S$ and a subjective utility function $u:X \rightarrow \mathbb{R} $, an action $a^*: S \rightarrow X $ is SEU-maximizing if $a^* \in \text{argmax}_{a_i \in A} \bigint_S u(a^*(s))dP(s) $. \\

The subjectivity of preferences plays a very important
role for the present work. From an agent's standpoint, the amount $x$ of, say, money achieved does not necessarily
coincide with the amount of subjective utility attained $u(x)$. 
Historically,  the first to notice and formally argue for this difference
between objective utility (goods, money, etc.) and subjective utility (preferences) was
D. Bernoulli with the famous St. Petersburg paradox in 1738. Moreover, Bernoulli introduced the idea that a good representation of
subjective utility can be carried out in terms of an increasing concave function of objective
utility. This can be considered the birth of modern decision theory.

Empirically, many violations of Savage's (and vNM's) axioms have been consistently observed in behavioral economics. Most of the attention has been focused on the Independence Axiom and the Sure-thing Principle. Allais' paradox and Ellsberg's paradox are examples of violations of these two axioms, and they are at the origin of Prospect theory [cite] and ambiguity in decision theory [cite].

The main idea behind Ellesberg's paradox and the literature on ambiguity is that DM treats differently objective and subjective uncertainty. If the probability measure on the possible states of the world is not given a priori and not ''known'', then observed preferences leave no room to represent  DM's beliefs via a probability measure on the possible states. The most common solution is then to represent DM's beliefs with a set of probability measures instead of a single one.

Prospect theory was introduced by Kahnemann and Tversky in
their famous paper from 1979. By means of experimental results they noticed that DM apparently acts as a risk averse agent with respect to gains and as a risk loving agent
with respect to losses. Consequently, they conjectured that the subjective utility is better
representable by an S-shaped utility function, concave in the positive domain of gains and
convex in the negative domain of losses.

More recently, other possible definitions of the subjective utility function have been proposed
and justified from a game theoretical point of view (e.g. Fehr and Schmidt, Charness and Rabin,
etc.). The central goal of these studies is to explain/describe the empirical behavior in social dilemmas
(e.g., Prisoner's dilemma, public goods game, etc.). In the same spirit of Prospect theory, by means of different subjective utility
functions it is possible to describe/match the behavior observed in game theoretical experiments and to
explain it as rational (in the sense of maximizing the subjective utility).
With this approach every behavior, no matter how far from the maximization of
objective utility, can be deemed rational if we can only find a subjective utility that
justifies that behavior as maximizing the subjective utility. The problem here is that it seems possible to consider rational everything
that is observed, given the adequate subjective utility function. The notion
of rationality almost collapses onto the observed behavior.

On the other side, other economists (Alchian, Friedman) seem to link the notion of rational choice
more tightly to the maximization of objective utility. In their classic works they argue that
profit maximization is a reasonable assumption for characterizing outcomes in a competitive
market because only firms behaving in a manner that is consistent with profit maximization will
survive in the long run. If the previous accounts of rationality looked more descriptive, this
definitely tastes more normative. In this sense, a normative notion of rationality would be tightly
related to (and almost coincide with) the maximization of objective utility, which is supposed
to drive the evolutionary process. It follows then that being normative in this sense points to an identification of rationality with maximization of objective utility.

@@

However, it has been argued that in a good theory of choice and decision making the normative and the descriptive components should not come apart. For example, as Douglas Maclean [cite] writes:

\begin{quote}
The history of the subject reflects strong reasons why the theory of choice should have this dual character. [...] \\
A normative theory of choice gains credibility from its predictive and explanatory power, for although empirical success itself may not constitute much of a normative argument, the theory of rational behavior is less likely to be found objectionable if rationality turns out to be widespread. [...] But if, on the other hand, the normative theory turns out to be incompatible with the best established behavioral theory, then the verification of the normative theory will depend entirely on the intrinsic plausibility of its axioms, and that appeal must be quite strong, strong enough to make us overcome our charitable reluctance to conclude that our actual behavior is systematically irrational.
\end{quote}

There is another violation of Savage's axioms that is particularly relevant for the present work: transitivity of preferences. The first of Savage's axioms states that the preference relation is a (total) preorder, i.e., reflexive and transitive. Moreover, the preorder axiom is often referred to as the rationality axiom [cite], as to suggest that it is somehow the core of rational preferences. Violations of transitivity have been observed and studied in experimental literature [citeTversky69,etc], but can also be supported from a normative and theoretical point of view. \\
In what follows, an important focus of our work will be the role of regret in decision making. Regret theory predicts the violation of both transitivity and Independence of Irrelevant Alternatives (IIA). [CONNECTING TO Fawcett\&al: this is IIA, not independence]

@@

Regret theory has been introduced in decision theory by Savage [] and Niherens [], and later developed by Bell [1982], Loomes and Sugden [1982], and Fishburn [1982] independently. Lately, Halpern and Pass [cite] show how the use of regret minimization can solve game theoretical puzzles (like the Centipede and the Traveller's dilemma) in a way that is closer to everyday intuition and empirical data. In this paper the notion of regret is defined as in [HalPass12]:

\vspace{.5cm}

@@

EXAMPLE 1: \textit{Coordination Game.}

Let us consider the coordination game in Figure \ref{coordgame1}.

\begin{figure}
\begin{center}%
\begin{tabular}{|c|c|c|}
\hline 
 & I & II\tabularnewline
\hline 
\hline 
I & 1;1 & 0;0\tabularnewline
\hline 
II & 0;0 & 2;2\tabularnewline
\hline 
\end{tabular}\end{center}

\protect\caption{A coordination game}
\label{coordgame1}
\end{figure}

It is easy to see that $\text{reg}(I)=2$: when the other player chooses action $I$ the regret of $I$ is $0$ since $I$ is a best reply to $I$, and when the other player chooses $II$ the regret of $I$ is $2$ since $\pi_i(II,II)-\pi_i(I,II)=2$. With the same reasoning, $\text{reg}(II)=1$. Two regret minimizing players would then choose action $II$, each getting a payoff of $2$. In this case regret minimization selects the Pareto-dominant Nash equilibrium. $ \medsquare $

\vspace{.5cm}

EXAMPLE 2: \textit{The Traveller's Dilemma.} (HalpernPass12)

Let us now consider the Traveller's Dilemma case. The story behind the game goes as follows. There are two travellers, Ann and Bob, who lost their luggages. Ann and Bob had exactly the same luggage and they had insured the luggages with the same insurance company. The insurance policy is that the two travellers have to separately claim a certain amount between 2\$ and 100\$ for the reimbursement, and if they both claim the same amount then they will be reimbursed by that amount. Otherwise, if one of them claims more than the other, then the one who claimed the higher amount will get the lower amount minus 2\$, while the one who claimed the lower amount will get what (s)he claimed plus 2\$. The game is symmetric, and the payoff function for both players is:\\
$\pi_i(a_i,a_{-i}) =  \begin{cases} a_i \text{ \hspace{1cm} if } a_i = a_{-i} \\ a_i + 2 \text{ \hspace{.4cm} if } a_i < a_{-i} \\ a_{-i} -2 \text{ \hspace{.25cm} if } a_{-i} < a_i \end{cases}$ \\
By a simple computation we have:  $\forall a_i \in \lbrace 96\$, ..., 100\$ \rbrace \text{, } \text{reg}(a_i)=3 $ and $\forall a_i \in \lbrace 2\$, ..., 95\$ \rbrace \text{, } \text{reg}(a_i)>3$. If we now eliminate all the actions that do not minimize regret in the first place and iterate the regret minimization only on the action set $\lbrace 96\$, ..., 100\$ \rbrace$, we get that 97\$ is the action that minimize regret. Indeed, $\text{reg}(97)=2$ and $\forall a_i \in \lbrace 96\$, 98\$, 99\$, 100\$ \rbrace \text{, } \text{reg}(a_i)=3 $. Hence, one step of regret minimizing reasoning eliminates all the actions smaller than 96\$, and if we iterate the process once more the unique outcome is 97\$. Finally, it is worth noticing that the only rationalizable equilibrium of the game corresponds to the outcome $(2,2)$, that is consequently the only Nash equilibrium, the only perfect equilibrium, and the only sequential equilibrium of the game. $ \medsquare $\\

The meta-game perspective that we introduce in this paper can shed more light on the duality of descriptivism and normativism, and opens
some space in between the two extrema. 
By keeping the evolutionary approach we want to retain a
normative perspective on rationality, but our results show that (under certain conditions)
evolution does not necessarily lead to identifying rationality with maximization of objective
utility, contrary to the intuition of Alchian [cite] and Friedman [cite].

@@

The justification for this choice is both historical and behavioral. Among
the possible competitors of utility maximization, regret has always been in a prominent
position, both from a descriptive and normative viewpoint. Indeed, Bell82 writes:
\begin{quote}
  By explicitly incorporating regret, expected utility theory not only becomes a better
  descriptive predictor but also may become a more convincing guide for prescribing behavior to
  decision makers.
\end{quote}
In the same year, Graham Loomes and Robert Sugden \citet{LoomesSugden1982:Regret-Theory:-} show that a regret-based theory of choice can match
and explain all the main experimental results of prospect theory, and claim that
\begin{quote} [...] we shall challenge the idea that the conventional axioms [of decision
  theory] constitute the only acceptable basis for rational choice under uncertainty. We shall
  argue that it is no less rational to act in accordance with regret theory, and that
  conventional expected utility theory therefore represents an unnecessarily restrictive notion
  of rationality. \hfill \citep{LoomesSugden1982:Regret-Theory:-}
\end{quote}
